Interpretable Machine Learning

### 2019-01-16 chunks/story-wines.Rmd
adds intuition shapley
polish slides

- ~6500 red and white Portuguese "Vinho Verde" wines
- Features: Physiochemical properties
- Quality assessed by blind tasting, from 0 (very bad) to 10 (excellent)

<font size="2">P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.
Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.</font>

@@

@@

## Compare Models with 10x CV {.emphasizedabit}

- Linear regression
- Decision tree
- Random forest

@@

## Compare Models with 10x CV {.emphasizedabit}

- Linear regression (`r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.lm"])`)
- Decision tree (`r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.rpart"])`)
- Random forest (`r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.ranger"])`)

=> Random forest has lowest Mean Absolute Error

@@

```{r check-model}

preds = pred$predict(wine)
preds$actual = wine$quality
ggplot(preds, aes(x = actual, y = .prediction, group = actual)) + 
  geom_violin(aes(fill = ..n..)) +
  scale_x_continuous("Actual quality", 
    labels = 1:10, breaks = 1:10) + 
  scale_y_continuous("Predicted quality", labels = 1:10, breaks = 1:10) + 
  scale_fill_gradient(low = "white", high = "darkblue", guide = "none")
```

@@

@@

@@

## How do features affect predictions?
```{r}
knitr::include_graphics("../images/ale-intuition-1.jpg")
```

## How do features affect predictions?
```{r}
knitr::include_graphics("../images/ale-intuition-2.jpg")
```

## How do features affect predictions?
```{r}
knitr::include_graphics("../images/ale-intuition-3.jpg")
```

## How do features affect predictions?
```{r}
knitr::include_graphics("../images/ale-intuition-4.jpg")
```

## How do features affect predictions?
```{r}
knitr::include_graphics("../images/ale-intuition-5.jpg")
```

## Effect of Alchohol
```{r}
eff = FeatureEffect$new(pred, "alcohol", method = 'ale')
plot(eff)
```

## Effect of Volatile Aciditys
```{r}
eff = FeatureEffect$new(pred, "volatile.acidity", method = 'ale')
plot(eff)
```

@@

## Interaction: Alcohol + volatile acidity

```{r}
inter = FeatureEffect$new(pred, feature = c("alcohol", "volatile.acidity"))
plot(inter, show.data = TRUE)
```

@@

## Rule of thumb for wine quality? {.center data-background=../images/global-surrogate.png data-background-size=contain}

## Rule of thumb for wine quality? 
```{r surrogate}
library(partykit)
tree = TreeSurrogate$new(pred, maxdepth  = 2)
node_inner2 = node_inner(tree$tree, pval = FALSE, id=FALSE)
node_terminal2 = node_boxplot(tree$tree, id = FALSE)
plot(tree$tree, inner_panel = node_inner2, terminal_panel = node_terminal2)

@@

## Explain individual prediction (e.g. worst wine)

@@

min_pred = min(predictions)
worst = wine[which(predictions == min_pred),]
```

@@

TODO: Slide to explain Shapley value

@@

@@

## Counterfactual explanation

@@

TODO: Image

>>>>

- 6500 red and white Portuguese "Vinho Verde" wines
- Features: Physicochemical properties
- Quality assessed by blind tasting, from 0 (very bad) to 10 (excellent)

```{r, out.width= "30%"}
knitr::include_graphics("../images/judges.png")
```

<font size="2">
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.
Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.</font>

@@

## How can we solve this problem? {.emphasizedabit}

- Hand-written rules
- Statistical Model
- Machine Learning

```{r, child="../chunks/what-is-ml.Rmd"}
```

## 
```{r}
knitr::include_graphics("../images/checkmark.png")
```

@@

## Random Forest

![](../images/random-forest.jpg)

<font size="2">
Image: 
http://www.hallwaymathlete.com/2016/05/introduction-to-machine-learning-with.html
</font>

@@

## Train Random Forest to Predict Quality {.emphasizedabit}

Mean absolute error on test data (cross-validated): `r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.ranger"])`

@@

```{r check-model}
preds = pred$predict(wine)
preds$actual = wine$quality
ggplot(preds, aes(x = actual, y = .prediction, group = actual)) + 
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  geom_violin() +
  scale_x_continuous("Actual quality", 
    labels = 1:10, breaks = 1:10, limit = c(2.5,9.5)) + 
  scale_y_continuous("Predicted quality", labels = 1:10, breaks = 1:10, limits = c(2.5,9.5)) 
#scale_fill_gradient(low = "white", high = "darkblue") + 
```

@@

## We want to know:

- Which wine properties are the most predictive for quality?
- How does a property affect the predicted wine quality?
- Can we extract a "Rule of Thumb" from the black box?
- Why did a wine get a certain prediction?
- How do we have to change a wine to achieve a different prediction?

@@

## Permutation Feature Importance

@@

## Accumulated Local Effects

```{r,  out.width = '78%'}
knitr::include_graphics("../images/ale-intuition-1.jpg")
```

## Accumulated Local Effects

```{r,  out.width = '78%'}
knitr::include_graphics("../images/ale-intuition-2.jpg")
```

## Accumulated Local Effects

```{r,  out.width = '78%'}
knitr::include_graphics("../images/ale-intuition-3.jpg")
```

## Accumulated Local Effects
```{r,  out.width = '78%'}
knitr::include_graphics("../images/ale-intuition-4.jpg")
```

## Accumulated Local Effects
```{r,  out.width = '78%'}
knitr::include_graphics("../images/ale-intuition-5.jpg")
```

## Effect of Alcohol
```{r}
eff = FeatureEffect$new(pred, "alcohol", method = 'ale')
plot(eff)
```

## Effect of Volatile Acidity
```{r}
eff = FeatureEffect$new(pred, "volatile.acidity", method = 'ale')
plot(eff)
```

@@

@@

## Surrogate Model {.center data-background=../images/global-surrogate.png data-background-size=contain}

## Surrogate Model
```{r surrogate}
library(partykit)
tree = TreeSurrogate$new(pred, maxdepth  = 2)
node_inner2 = node_inner(tree$tree, pval = FALSE, id=FALSE)
node_terminal2 = node_boxplot(tree$tree, id = FALSE)
plot(tree$tree, inner_panel = node_inner2, terminal_panel = node_terminal2)

@@

## Explain individual predictions

@@

min_pred = min(predictions)
max_pred = max(predictions)
worst = wine[which(predictions == min_pred),]
best = wine[which(predictions == max_pred),]
```

```{r}
knitr::include_graphics("../images/shapley-intuition-1.png")
```

## Explain best wine

```{r}
shap = Shapley$new(pred, x.interest = best)
plot(shap) + 
  scale_x_discrete("") + 
  scale_y_continuous("Feature contribution")
```

@@

## Explain worst wine

@@

## Improve worst wine?

@@

## Improve worst wine?

@@
